Improving Outcome Predictions for Patients Receiving Mechanical Circulatory Support by Optimizing Imputation of Missing Values

### 2020-10-13 doc_arXiv/doc_arXiv.tex
misc changes before using outcomes

Previous research has thoroughly investigated statistical inference in
the context of missing data. In this setting, analysts often create
multiple imputed datasets, replicate an analysis in each of them, and
then pool their results to obtain valid test statistics for hypothesis
testing \cite{rubin2004multiple}. Imputation strategies that create a
single imputed dataset (\textit{e.g., }imputation to the mean) have been
shown to increase type I errors (\textit{i.e., }rejecting a true null
hypothesis) for inferential statistics by artificially reducing the
variance of observed data and ignoring the uncertainty attributed to
missing values \cite{van2018flexible}. When very few data are missing,
analysts may apply listwise deletion, which removes any observation with
at least one missing value and produces a dataset with fewer
observations and no missing values.

@@

@@

The random forest framework can be used for imputation as well as
prediction. In the current analysis, we implement \texttt{miceRanger},
an R package similar to \texttt{missForest} that implements MICE with
random forests by implementing PMM \cite{missForest}.
\texttt{miceRanger} follows an iterative procedure where one random
forest is fit to each variable in a dataset, separately, and then used
to impute missing values in that variable. The procedure may repeat
until estimated generalization error of the random forest collection
stabilizes. When imputing a single dataset with \texttt{miceRanger}, we
did not use PMM and instead imputed missing values with a prediction.
When imputing multiple datasets, we implemented PMM with 10 donor
candidates.

@@

MIA adds another observed category to nominal variables: ``Missing''.
For numeric variables, MIA creates two columns: one where missing values
are imputed with positive infinity and the other with negative infinity.
When a decision tree uses a finite cut-point to split a given numeric
variable, it will assess the cut-point using both the positive and
negative infinite imputed columns, and utilize whichever column provides
the best split of the current data. This procedure translates to sending
all missing values to the left or to the right when forming two new
nodes of the decision tree, using whichever direction results in a
better split. In the current analysis, we used the MIA strategy only for
ensembles of boosted decision trees, as it is a well known and
frequently used strategy for these models
\cite{twala2009empirical, ding2010investigation}.

\paragraph{Surrogate splitting}

When decision trees search for a variable and cut-point to use for
growing two new nodes in a decision tree, surrogate splitting will
temporarily ignore missing values while searching for an initial split.
Once an initial split is chosen, surrogate splitting searches for a
split on another variable with observed values for the observations with
missing values on the initial splitting variable and creates a surrogate
split based on the new variable that is similar to the original split.
In the current analysis, we used a maximum of three surrogate splits and
sent any remaining missing values to the daughter node with higher
sample size. Additionally, we only used surrogate splitting for
conditional inference forests, as this strategy is well known and
frequently used for these models.

@@

Each missing data strategy poses different assumptions regarding the
data and the mechanisms that lead to missing values in the data.
Imputation using Bayesian regression makes the same distributional
assumptions as the Bayesian models that are applied. The
\texttt{miceRanger} algorithm does not make any formal distributional
assumptions, as random forests are non-parametric and can thus handle
skewed and multi-modal data as well as categorical data that are ordinal
or non-ordinal. Hot deck and KNN imputation also do not make
distributional assumptions, but implicitly assume that a missing value
for a given observation can be approximated by aggregating observed
values from the \(k\) most similar observations. PMM makes a similar
implicit assumption, but is slightly more robust to skewed or
multi-model data because imputed values are sampled directly from
observed ones. MIA operates based on an implicit assumption that
missingness itself is informative and surrogate splitting presumes that
variables with missing values are related to other observed variables
used by the decision tree. It is difficult to validate these assumptions
in applied settings and also likely that downstream models will perform
poorly if an imputation technique's assumptions are invalid.

@@

We applied three modeling strategies after imputing missing values:

\begin{itemize}
\tightlist
\item
  Cox proportional hazards (PH) model with forward stepwise variable
  selection
\item
  gradient boosted decision trees (hereafter referred to as `boosting')
\item
  Conditional inference forest (CIF, an extension of random forests)
\end{itemize}

A thorough description of stepwise variable selection, boosting, and
random forests can be found in Sections 6.1.2, 8.2.2, and 8.2.3,
respectively, of \emph{Introduction to Statistical Learning}
\cite{james2013introduction}. Sir David Cox's PH model is one of the
most frequently applied methods for the analysis of right-censored
time-to-event outcomes \cite{kleinbaum2010survival}. According to the PH
assumption, the effect of a unit increase in a predictor is
multiplicative with respect to a baseline hazard function. Boosting
grows a sequence of decision trees, each using information from the
previous trees in an attempt to correct their errors
\cite{friedman2001greedy, chen2016xgboost}. In contrast, random forests
grow an ensemble of de-correlated decision trees, where each tree is
grown using a bootstrapped replicate of the original training data
\cite{Breiman2001}. The CIF obtains unbiased recursive partitioning by
applying early-stopping and permutation tests and is one of several
methods that extends random forests to right-censored time-to-event data
\cite{hothorn2006survival, strobl2007bias, strobl2008conditional, ishwaran2008random, jaeger2019oblique}.

@@

To assess the prognostic value of each missing data strategy, we
internally validated a total of FILL IN modeling algorithm based on
combinations of imputation strategies and modeling strategies described
in Sections \ref{subsec:imputation} and \ref{subsec:modeling}. For
convenience, we use the term `modeling algorithm' to denote the
combination of a missing data strategy and a modeling strategy
(\textit{e.g., }imputation using mean/mode values followed by fitting
the PH model with stepwise variable selection) \cite{kuhn2013applied}.
We conducted internal validation using 150 replicates of Monte-Carlo
cross validation (MCCV), a resampling technique for internal validation
(Figure \ref{fig:monte_carlo_cv}).

@@

>>>>

Previous research has thoroughly investigated statistical inference in
the context of missing data. In this setting, analysts often create
multiple imputed datasets, replicate an analysis in each of them, and
then pool their results to obtain valid test statistics for hypothesis
testing \cite{rubin2004multiple}. Imputation strategies that create a
single imputed dataset (\textit{e.g., }imputation to the mean) have been
shown to increase type I errors (\textit{i.e., }rejecting a true null
hypothesis) for inferential statistics by artificially reducing the
variance of observed data and ignoring the uncertainty attributed to
missing values \cite{van2018flexible}. To ensure valid inference,
imputation models should leverage outcome variables to predict missing
predictors \cite{sterne2009multiple}. When very few data are missing,
analysts may apply listwise deletion, \textit{i.e., }removing any
observation with at least one missing value. However, listwise deletion
can easily lead to biased inference \cite{van2020rebutting}.

@@

Though the outcome variable may be useful for imputing predictor values,
using outcome variables for this purpose in supervised learning projects
can lead to issues in clinical implementation. For instance, suppose we
seek to predict an outcome \(Y\) for a patient in the clinical setting.
If the patient is missing information for a predictor \(X\) and our only
method to impute missing values of \(X\) leverages the observed value of
\(Y\), how should we impute \(X\)? If we do not already know \(Y\),
which would seem likely for clinical application of a prediction model,
then we cannot impute \(X\). On the other hand, if we already know
\(Y\), then we obviously do not need to predict it. Due to these
practical considerations and because INTERMACS is often leveraged to
create prediction models for clinical practice \cite{thomas2014pre}, we
did not leverage outcome variables (\textit{i.e., }time until
transplant, death, or last contact) to impute missing values of
predictors in the current analysis.

@@

Random forests grow an ensemble of de-correlated decision trees, where
each tree is grown using a bootstrapped replicate of the original
training data
\cite{Breiman2001, hothorn2006survival, strobl2007bias, strobl2008conditional, ishwaran2008random, jaeger2019oblique}.
A particularly helpful feature of random forests is their ability to
estimate testing error by aggregating each decision tree's prediction
error on data outside of their bootstrapped sample
(\textit{i.e., }out-of-bag error). In the current analysis, we conduct
MICE using one random forest to impute each variable, separately. For
each imputed dataset, we allowed random forests to be re-fitted until
out-of-bag error stabilized or a maximum number of iterations was
completed. When imputing a single dataset, we used 250 trees per forest
and a maximum of 10 iterations. When imputing multiple datasets, we used
50 trees per forest and applied PMM using the random forest's predicted
values to impute missing data by sampling one value from a pool of 10
potential donors.

@@

MIA is a technique that uses missing status as a predictor rather than
explicitly imputing missing values
\cite{twala2009empirical, ding2010investigation}. MIA adds another
category to nominal variables: ``Missing''. For numeric variables, MIA
creates two columns: one where missing values are imputed with positive
infinity and the other with negative infinity. Since PH models are not
compatible with infinite values, we only use MIA in boosting models.
When a decision tree uses a finite cut-point to split a given numeric
variable, it will assess the cut-point using both the positive and
negative infinite imputed columns, and utilize whichever column provides
the best split of the current data. This procedure translates to sending
all missing values to the left or to the right when forming two new
nodes of the decision tree, using whichever direction results in a
better split.

@@

Each missing data strategy poses different assumptions regarding the
data and the mechanisms that lead to missing values in the data.
Imputation using Bayesian regression makes the same distributional
assumptions as the Bayesian models that are applied. The
\texttt{miceRanger} algorithm does not make any formal distributional
assumptions, as random forests are non-parametric and can thus handle
skewed and multi-modal data as well as categorical data that are ordinal
or non-ordinal. Hot deck and KNN imputation also do not make
distributional assumptions, but implicitly assume that a missing value
for a given observation can be approximated by aggregating observed
values from the \(k\) most similar observations. PMM makes a similar
implicit assumption, but is slightly more robust to skewed or
multi-modal data because imputed values are sampled directly from
observed ones. MIA operates based on an implicit assumption that
missingness itself is informative. It is difficult to validate these
assumptions in applied settings and also likely that downstream models
will perform poorly if an imputation technique's assumptions are
invalid.

\hypertarget{evaluating-imputation-accuracy}{%
\subsection{Evaluating imputation
accuracy}\label{evaluating-imputation-accuracy}}

\label{subsec:imputation_accuracy}

Imputation accuracy was computed for each numerical and nominal
variable, separately. Numeric variable imputation accuracy was measured
using a re-scaled mean-squared error:
\[1 - \frac{\textrm{MSE}(\textrm{current imputation method})}{\textrm{MSE}(\textrm{imputation to the mean})}.\]
This score is greater than 0 if
\(\textrm{MSE}(\textrm{current imputation method})\) is smaller than
\(\textrm{MSE}(\textrm{imputation to the mean})\), equal to 0 if the two
MSEs are equal, and less than 0 if
\(\textrm{MSE}(\textrm{current imputation method})\) is greater than
\(\textrm{MSE}(\textrm{imputation to the mean})\). Nominal variable
imputation accuracy was measured using a re-scaled classification
accuracy:
\[1 - \frac{\textrm{Classification error}(\textrm{current imputation method})}{\textrm{Classification error}(\textrm{imputation to the mean})}.\]
This score is greater than 0 if the classification error of the current
imputation method is less than that of imputation to the mean, equal to
0 if the two classification errors are equal, and less than 0 of the
classification error of the current imputation method is worse than
imputation to the mean. These numeric and nominal scores are analogous
to the more well known \(R^2\) and Kappa statistics, respectively, but
are modified slightly in the current analysis so that imputation to the
mean will always have a score of 0. This modification makes it easier to
compare the accuracy of each imputation strategy directly with
imputation to the mean.

@@

We applied two modeling strategies after imputing missing values:

\begin{itemize}
\tightlist
\item
  Cox proportional hazards (PH) model with forward stepwise variable
  selection
\item
  Gradient boosted decision trees (hereafter referred to as `boosting')
\end{itemize}

A thorough description of stepwise variable selection and boosting can
be found in Sections 6.1.2 and 8.2.2, respectively, of
\emph{Introduction to Statistical Learning}
\cite{james2013introduction}. Sir David Cox's PH model is one of the
most frequently applied methods for the analysis of right-censored
time-to-event outcomes \cite{kleinbaum2010survival}. According to the PH
assumption, the effect of a unit increase in a predictor is
multiplicative with respect to a baseline hazard function. Boosting
grows a sequence of decision trees, each using information from the
previous trees in an attempt to correct their errors
\cite{friedman2001greedy, chen2016xgboost}.

@@

To assess the prognostic value of each missing data strategy, we
internally validated a total of FILL IN modeling algorithm based on
combinations of imputation strategies and modeling strategies described
in Sections \ref{subsec:imputation} and \ref{subsec:modeling}. For
convenience, we use the term `modeling algorithm' to denote the
combination of a missing data strategy and a modeling strategy
(\textit{e.g., }imputation using mean/mode values followed by fitting
the PH model with stepwise variable selection) \cite{kuhn2013applied}.
We conducted internal validation using 200 replicates of Monte-Carlo
cross validation (MCCV), a resampling technique for internal validation
(Figure \ref{fig:monte_carlo_cv}).

@@

\hypertarget{statistical-analysis}{%
\subsection{Statistical analysis}\label{statistical-analysis}}

Participant characteristics were tabulated for the overall population
and stratified by event status. Continuous and categorical variables
were summarized as mean (standard deviation) and percent, respectively.
Imputation accuracy was aggregated for all numeric and nominal variables
to create two overall scores for each imputed dataset, and the
distribution of these scores was tabulated for each imputation strategy.

\hypertarget{computational-details}{%
\subsection{Computational details}\label{computational-details}}

\label{subsec:computing}

implement \texttt{miceRanger}, an R package similar to
\texttt{missForest} that implements MICE with random forests by
implementing PMM \cite{missForest}. \texttt{miceRanger} follows an
iterative procedure where one random forest is fit to each variable in a
dataset, separately, and then used to impute missing values in that
variable. The procedure may repeat until estimated generalization error
of the random forest collection stabilizes.

\hypertarget{results}{%
\section{Results}\label{results}}

\paragraph{Patient Characteristics}

During the first 6 months after receiving MCS, the number (\% of overall
population) of deaths, transplants, and loss-to-follow-up events was
1,897 (13\%), 1,067 (7.2\%), and 981 (6.7\%), respectively. The mean
(standard deviation) age of patients was 57 (13) years, 66\% of patients
identified as white and 79\% were male (Table
\ref{tbl_characteristics}).

\paragraph{Imputation accuracy}

\clearpage
\begin{table}
\caption{Participant characteristics stratified by event status.}
\label{tbl_characteristics}

\begin{tabular}{lccccc}
\toprule
Characteristic & Overall & Censored & Cessation of support & Dead & Transplant\\
\midrule
No. of patients & 14,738 & 6,308 & 310 & 4,204 & 3,916\\
Follow up time, months & 18 (16) & 25 (18) & 16 (11) & 13 (14) & 13 (10)\\
Age, years & 57 (13) & 57 (13) & 48 (14) & 61 (12) & 53 (12)\\
Sex &  &  &  &  & \\
\hspace{1em}Female & 21 & 22 & 41 & 22 & 19\\
\addlinespace
\hspace{1em}Male & 79 & 78 & 59 & 78 & 81\\
Race &  &  &  &  & \\
\hspace{1em}White & 66 & 63 & 64 & 71 & 67\\
\hspace{1em}Black & 25 & 28 & 27 & 21 & 23\\
\hspace{1em}Other & 9.2 & 9.6 & 9.4 & 7.4 & 10\\
\addlinespace
Body mass index & 29 (7) & 29 (7) & 29 (8) & 29 (7) & 28 (6)\\
Device strategy &  &  &  &  & \\
\hspace{1em}Other & 0.6 & 0.6 & 2.6 & 0.5 & 0.6\\
\hspace{1em}Bridge to transplant & 52 & 43 & 51 & 40 & 81\\
\hspace{1em}Destination therapy & 47 & 56 & 46 & 60 & 18\\
\addlinespace
Device type &  &  &  &  & \\
\hspace{1em}Bi-ventricular assistance device & 3.5 & 2.1 & 0.3 & 5.8 & 3.4\\
\hspace{1em}Left-ventricular assistance device & 97 & 98 & 100 & 94 & 97\\
\bottomrule
\multicolumn{6}{l}{\rule{0pt}{1em}\textit{Note: }}\\
\multicolumn{6}{l}{\rule{0pt}{1em}Table values are mean (standard deviation) or \%}\\
\end{tabular}

\end{table}

\clearpage

\begin{table} 
 \caption{Accuracy of strategies to impute artificial missing data} 
\label{tbl_impute_accuracy}

\begin{tabular}{lcccc}
\toprule
& \multicolumn{2}{c}{Nominal variables} & \multicolumn{2}{c}{Numeric variables} \\ 
 \cmidrule(lr){2-3}\cmidrule(lr){4-5}
 & Single imputation & Multiple imputation & Single imputation & Multiple imputation \\ 
\midrule
\multicolumn{1}{l}{+15\% additional missing data} \\ 
\midrule
Imputation to the mean & 0 (reference) & --- & 0 (reference) & --- \\ 
Hot deck & -0.24 (-0.34, -0.16) & -0.25 (-0.37, -0.18) & -0.35 (-0.53, -0.29) & -0.42 (-0.54, -0.36) \\ 
K-nearest-neighbors & 0.02 (-0.05, 0.15) & -0.19 (-0.29, -0.08) & 0.01 (-0.01, 0.03) & -0.33 (-0.42, -0.28) \\ 
Predictive mean matching & -0.15 (-0.34, 0.00) & -0.15 (-0.35, 0.00) & -0.25 (-0.40, -0.20) & -0.26 (-0.35, -0.21) \\ 
Random forest & -0.15 (-0.26, -0.02) & -0.15 (-0.26, -0.03) & 0.06 (0.02, 0.08) & 0.05 (0.02, 0.07) \\ 
Bayesian regression & -0.14 (-0.26, 0.01) & -0.14 (-0.27, 0.01) & -0.30 (-0.41, -0.23) & -0.30 (-0.41, -0.23) \\ 
\midrule
\multicolumn{1}{l}{+30\% additional missing data} \\ 
\midrule
Imputation to the mean & 0 (reference) & --- & 0 (reference) & --- \\ 
Hot deck & -0.25 (-0.34, -0.18) & -0.26 (-0.36, -0.18) & -0.34 (-0.46, -0.30) & -0.40 (-0.53, -0.36) \\ 
K-nearest-neighbors & -0.04 (-0.09, 0.03) & -0.24 (-0.34, -0.16) & -0.02 (-0.03, -0.01) & -0.36 (-0.44, -0.32) \\ 
Predictive mean matching & -0.18 (-0.38, -0.04) & -0.18 (-0.37, -0.04) & -0.29 (-0.46, -0.24) & -0.29 (-0.40, -0.25) \\ 
Random forest & -0.18 (-0.27, -0.07) & -0.18 (-0.28, -0.07) & -0.02 (-0.05, 0.01) & -0.01 (-0.04, 0.01) \\ 
Bayesian regression & -0.17 (-0.29, -0.04) & -0.17 (-0.29, -0.05) & -0.31 (-0.43, -0.27) & -0.31 (-0.43, -0.27) \\ 
\bottomrule
\end{tabular} 
 \end{table}
